---
title: "Mitch Ryan, Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
In this project we look at some data taken from wearable fitness devices. The data contains 3 variables,
steps-the amount of steps taken in a 5 min interval
date-the date the observation was taken
interval-the 5 minute interval in question

## Loading and preprocessing the data
This code will load in our dataset from the url, and unzip it.
The last line removes scientific notation, which will be necessary for our analysis
```{r, echo = TRUE}
# Download zip file if it doesn't already exist in the workspace
dataurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
if (!file.exists("activity.zip")) {
  download.file(dataurl, destfile = "activity.zip", method = "curl")
}
# Unzip data
datafile <- "activity.csv"
if (!file.exists(datafile)) {
  unzip(zipfile = "activity.zip")
}
# Load data
activity <- read.csv(datafile, header = TRUE)
options(scipen = 100)
```


## What is mean total number of steps taken per day?
```{r, echo = TRUE}
library(dplyr)
steps_date <- select(activity, steps, date, interval) %>%
  group_by(date) %>% summarize(total_steps = sum(steps, ra.rm = TRUE))
```
Quick line to make the NA values = 0
```{r, echo = TRUE}
steps_date[is.na(steps_date)] = 0
```
Now let's look at our new data frame, steps_date
```{r, echo = TRUE}
head(steps_date)
```

Now that we have our data the way we want it, we can find the mean and median,
and then make our histogram. 

First, let's find the mean and the median, because we want those on the plot

```{r, echo = TRUE}
mean_steps <- with(steps_date, mean(total_steps))
median_steps <- with(steps_date, median(total_steps))
```
To make the histogram we will use ggplot
```{r, echo = TRUE}
library(ggplot2)
a <- ggplot(steps_date, aes(total_steps))
a + geom_histogram(fill = 'blue') + labs(title = 'Total Step Frequency', x = 'Total Steps', y = 'Frequency') +
  geom_vline(aes(xintercept = mean_steps, color = 'mean'), linetype = 'dashed', size = 1) +  
  geom_vline(aes(xintercept = median_steps, color = 'median'), linetype = 'dashed',  size = 1) + 
  scale_color_manual(values = c(mean = 'red', median = 'yellow'))
```


The mean number of total steps is `r round(mean_steps, 2)` and is denoted by the red line.
The median number of total steps is `r round(median_steps, 2)` and is denoted by the yellow line.

## What is the average daily activity pattern?
```{r, echo = TRUE}
steps_interval <- select(activity, steps, date, interval) %>%
  group_by(interval) %>% summarize(average_steps = mean(steps, na.rm = TRUE))
```
Here is what our new data frame, steps_interval looks like
```{r, echo = TRUE}
head(steps_interval)
```

Let's calculate the maximum average steps, and the interval where that max occurs
```{r, echo = TRUE}
max_average_steps <- max(steps_interval$average_steps)
max_average_steps_interval <- steps_interval[
  steps_interval$average_steps == max_average_steps,]$interval
```

Now we make the plot
```{r, echo = TRUE}
b <- ggplot(steps_interval, aes(interval, average_steps))
b + geom_line(color = 'blue') + labs(title = 'Average Steps by Interval', x = 'Interval', y = 'Average Steps') + 
  geom_hline(aes(yintercept = max_average_steps, color = 'max'), linetype = 'dashed', size = 1) +
  geom_vline(aes(xintercept = max_average_steps_interval, color = 'interval'), linetype = 'dashed',  size = 1) +
  scale_color_manual(values = c(max = 'red', interval = 'green'))
```

## Imputing missing values
First, let's just see how many missing values there are
```{r, echo = TRUE}
na_amount <- nrow(activity[which(is.na(activity$steps)),])
```
There are `r na_amount` missing values in our activity dataset, which skews our data 
and makes it difficult to conduct analysis
So, let's see if we can fix that
```{r, echo = TRUE}
cleaned_activity <- activity
```
That line just made a copy of the dataset, now we are going to for-loop through 
our data, and replace every missing step value with the average steps from that interval
```{r, echo = TRUE}
for (i in 1:nrow(cleaned_activity)){
  if (is.na(cleaned_activity$steps[i])){
    int <- cleaned_activity$interval[i]
    avg <- steps_interval[steps_interval$interval==int,]$average_steps
    cleaned_activity$steps[i] <- avg
    
  }
}
```
Let's look at the new data frame
```{r, echo = TRUE}
head(cleaned_activity)
```

Now we can conduct the same analysis we did in the first part, but with our new cleaned data
```{r, echo = TRUE}
cleaned_steps_date <- select(cleaned_activity, steps, date, interval) %>%
  group_by(date) %>% summarize(total_steps = sum(steps, ra.rm = TRUE))
```
Finding the mean and median again:
```{r, echo = TRUE}
cleaned_mean_steps <- with(cleaned_steps_date, mean(total_steps))
cleaned_median_steps <- with(cleaned_steps_date, median(total_steps))
```

Make the histogram again
```{r, echo = TRUE}
c <- ggplot(cleaned_steps_date, aes(total_steps))
c + geom_histogram(fill = 'blue') + labs(title = 'Total Step Frequeny without NA values', x = 'Total Steps', y = 'Frequency') +
  geom_vline(aes(xintercept = cleaned_mean_steps, color = 'mean'), linetype = 'dashed', size = 1) +  
  geom_vline(aes(xintercept = cleaned_median_steps, color = 'median'), linetype = 'dashed',  size = 1) + 
  scale_color_manual(values = c(mean = 'red', median = 'yellow'))
```


In the new data, our mean and median are actually the same, `r round(cleaned_mean_steps, 2)`. 
This is denoted by the yellow dashed line. 

## Are there differences in activity patterns between weekdays and weekends?
Let's make add some new columns to the data frame, 
'weekday' which tells us the day of the week and
'weekday_type' which tells us weekday or weekend
```{r, echo = TRUE}
activity_weekday <- mutate(cleaned_activity, 
                    weekday = weekdays(as.Date(cleaned_activity$date)),
                    weekday_type = as.factor(ifelse(weekday=="Saturday" | weekday=="Sunday", "weekend", "weekday")))
```
Let's look at that new dataset
```{r, echo = TRUE}
head(activity_weekday)
```

Now we get the average steps for weekend/weekday and interval
```{r, echo = TRUE}
summary_weekday <- select(activity_weekday, steps, date, interval, weekday_type) %>% 
  group_by(weekday_type, interval) %>% 
  summarize(average_steps = mean(steps, na.rm = TRUE))
```
And finally, making the plot

```{r, echo = TRUE}
d <- ggplot(summary_weekday, aes(interval, average_steps))
d + geom_line(color = 'red') + facet_grid(weekday_type ~.) +
  labs(title = 'Average Steps over Interval by Day Type', x = 'Average Steps', y = 'Interval')